---
title: "Class Size: Wisdom and Justice"
format:
  html: default
execute:
  echo: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)

star <- read_csv("data/class-size.csv", show_col_types = FALSE) |>
  mutate(kind = recode(classtype,
                       `1` = "small",
                       `2` = "regular",
                       `3` = "regular with aid")) |>
  mutate(race = recode(race,
                       `1` = "White",
                       `2` = "Black",
                       `3` = "Others",
                       `4` = "Others",
                       `5` = "Others",
                       `6` = "Others")) |>
  select(g4math, kind, race, yearssmall)
```

## The Question

The Tennessee STAR experiment (Mosteller 1995) randomly assigned more than 6,000 students in about 80 Tennessee schools to small classes (13–17 students), regular classes (22–25), or regular classes with a teacher's aide, from kindergarten through third grade. Our data records each student's kindergarten class-type assignment, race, years spent in small classes across grades K through 3, and fourth-grade score on the math computation portion of the Stanford Achievement Test.

There are two scenarios. Focus on just one of them:

- Imagine that you are an elementary school principal in Chicago. You want to predict student performance. Fortunately, you have data today for your current students like the data available in the STAR project.

- Imagine you work for the Texas Department of Education. You want to understand the effect of small classes on student performance in Dallas. In Texas, there is data available like the data in the STAR project.

The principal asks: What will be the average fourth-grade math score of the students in my school next year, given their class type and background?

The Texas Department of Education asks: What is the average causal effect of assigning a Dallas student to a small class on their fourth-grade math score?

## Wisdom

### Preceptor Table

**What are the units, outcome, and covariates for your scenario's question?**

- Units:
K-3 grade students from Dallas.


- Outcomes:
Their fourth grade math scores with a regular class/small class/regular class with aides.


- Covariates:
Race, years spent in a small class, small/regular size.


- Treatment (if any):
If the student is assigned to a small class, regular class, or a regular class with teacher aides. (Causal)


- Quantity of interest:
The average causal effect (difference between the expected scores of the potential outcomes). (Causal)


Describe the key components of a Preceptor Table for your scenario; Use words like "units," "outcomes," and "covariates."
The preceptor table would have a column for the units (each 4th grade student), three columns for their fourth math score, and columns for the treatments.


### Exploratory Data Analysis

Create a plot that shows the distribution of `g4math`.

```{r}
#| label: g4math-distribution

ggplot(star, aes(x = g4math)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of 4th Grade Math Scores",
    x = "4th Grade Math Score",
    y = "Number of Students"
  )
```


Create a plot that shows the relationship between `g4math` and class type.

```{r}
#| label: g4math-vs-class-type

ggplot(star, aes(x = kind, y = g4math, fill = kind)) +
  geom_boxplot() +
  labs(
    title = "Fourth-Grade Math Scores by Class Type",
    x = "Class Type",
    y = "Math Score"
  ) +
  theme(legend.position = "none")
```




------------------------------------------------------------------------

## Justice

### Population Table

Describe the key components of a Population Table for your scenario.
Columns: Each student (units); small class score, regular class score, aides class score (outcomes); class size (treatment).


### Assumptions

Counter-example to the assumption of validity:
If the 4th grade math tests are not uniform between the students. The data is from 1985, the tests don't really match up.
Class sizes are subjective, they might not mean the same number of students per class.
Some students might have different amounts of years with aides, or another treatment.


Counter-example to the assumption of stability:
The introduction of technology might change teaching methods between then and now.


Counter-example to the assumption of representativeness:
The data doesn't really represent all the students in Texas, only Dallas.
If the data is not really representative of all the income brackets.
The Tennessee schools volunteered to participate in the data collection. A volunteered smaple can't represent everyone.


Counter-example to the assumption of unconfoundedness:
If the data only gave treatment to students who were currently doing poorly in their math class.


### Probability Family and Link Function

What probability family should we use?
Normal Distribution


Why should we use that probability family?
Because our model will be a continuous, linear regression.


What is the link function for our outcome?
A linear regression identity link.


------------------------------------------------------------------------
